---
title: 'Machine Learning 2019: Tree-Based Methods'
author: "Allen Zhu"
date: "10/28/2019"
output:
  html_document:
    df_print: paged
  pdf: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tree-Based Methods 

Decision tree is a type of supervised learning algorithm that can be used in both regression and classification problems. Tree-based methods works for both categorical and continuous input and output variables.

```{r load relevant libraries, include=FALSE}
library(tidyverse)
library(ISLR)
library(tree)
library(randomForest)
library(MASS)
library(gbm)
library(rpart)
library(rpart.plot)
```
## The Carseats Dataset 

400 Observations, 11 variables
Response Variable: Sales/High 

```{r The Carseats Dataset}
data("Carseats")
carseats = Carseats
head(carseats)

#convert quantitative variable Sales into a binary response 
High = ifelse(carseats$Sales<=8, "No", "Yes")
carseats = data.frame(carseats, High)

head(carseats)
```

## Classification Tree

Input variables (X) can be continuous or categorical.
Response variable (Y) is categorical (usually binary): in this case Sales/High.

```{r Classification Tree}
#set seed to make results reproducible 
set.seed(29)

#split data into train and test subset (250 and 150 respectively)
train = sample(1:nrow(carseats), 250)

#Fit train subset of data to model 
tree.carseats = tree(High~.-Sales, carseats, subset=train)
summary(tree.carseats)

#Visualize tree
plot(tree.carseats)
text(tree.carseats, pretty=0)

#each of the terminal nodes are labeled Yes or No. The variables and the value of the splitting choice are shown at each terminal node. 

#Use model on test set, predict class labels 
tree.pred = predict(tree.carseats, carseats[-train,], type="class")

#Misclassification table to evaluate error 
with(carseats[-train,], table(tree.pred, High))

#Calculate error by summing up the diagonals and dividing by number of total predictions
mc = (71 + 42) / 150
mc
```

## Pruning using cross-validation
Pruning is a method to cut back the tree to prevent over-fitting. 

```{r Pruning}
#cross-validation to prune the tree using cv.tree
cv.carseats = cv.tree(tree.carseats, FUN = prune.misclass)

#Sizes of the trees as they were pruned back, the deviances as the pruning proceeded, and cost complexity parameter used in the process.
cv.carseats

#Visualize 
plot(cv.carseats)

#Prune tree to a size of 12 
prune.carseats = prune.misclass(tree.carseats, best = 12)

#Visualize tree 
plot(prune.carseats)
text(prune.carseats, pretty=0)

#Evaluate on test set 
tree.pred = predict(prune.carseats, carseats[-train,], type="class")

#Misclassification 
with(carseats[-train,], table(tree.pred, High))

#Error 
mc_pruning = (66 + 41) / 150
mc_pruning

##pruning did not increase misclassification error by too much and resulted in a simpler tree!!
```
Pruning did not increase misclassification error by too much and resulted in a simpler tree!!

Decision trees suffer from high variance, meaning if you split the training data into 2 parts at random, and fit a decision tree to both halves, the results that you get could be very different.

Bagging and boosting are technique used to reduce the variance of your predictions.

## The Boston Housing Dataset 

506 Observations, 14 variables
Response Variable: medv (median value of owner-occupied homes for each suburb)

```{r The Boston Housing Dataset}
data("Boston")
boston = Boston
head(Boston)
```

## Bagging: Random Forest 

Bagging involves creating multiple copies of the original training dataset using the bootstrap, fitting a separate decision tree to each copy, and then combining all of the trees in order to create a single predictive model. Each tree is built on a bootstrapped dataset, independent of the other trees.

Random Forest: Each time a split in a tree is considered, a random sample of m predictors is chosen as split candidates from the full set of p predictors. The split is allowed to use only one of those m predictors.


```{r Random Forest}
#set seed for reproducibility 
set.seed(29)

#split into train and test sets (300 and 206 respectively)
train = sample(1:nrow(boston), 300)

#fit training subset of data to model 
rf.boston = randomForest(medv~., data = boston, subset = train)
rf.boston

#summary of rf.boston gives information about the number of trees, the mean squared residuals (MSR), and the percentage of variance explained

#No. of variables tried at each split: 4 
#Each time the tree comes to split a node, 4 variables would be selected at random, then the split would be confined to 1 of those 4 variables.

##Lets try a range of mtry (number of variables selected at random at each split)

oob.err = double(13)
test.err = double(13)

#In a loop of mtry from 1 to 13, you first fit the randomForest to the train dataset
for(mtry in 1:13){
  fit = randomForest(medv~., data = boston, subset=train, mtry=mtry, ntree = 350)
  oob.err[mtry] = fit$mse[350] ##extract Mean-squared-error 
  pred = predict(fit, boston[-train,]) #predict on test dataset
  test.err[mtry] = with(boston[-train,], mean( (medv-pred)^2 )) #compute test error
}

#Visualize 
matplot(1:mtry, cbind(test.err, oob.err), pch = 23, col = c("red", "blue"), type = "b", ylab="Mean Squared Error")
legend("topright", legend = c("OOB", "Test"), pch = 23, col = c("red", "blue"))
```

## Boosting 

Boosting is another approach to improve the predictions resulting from a decision tree. Trees are grown sequentially: each tree is grown using information from previously grown trees. Each tree is fitted on a modified version of the original dataset.


```{r Boosting}
#Gradient Boosting Model
boost.boston = gbm(medv~., data = boston[train,], distribution = "gaussian", n.trees = 10000, shrinkage = 0.01, interaction.depth = 4)

#Variable Importance Plot
summary(boost.boston)

#Visualize important variables of interest
plot(boost.boston,i="lstat")
plot(boost.boston,i="rm")

#Predict on test set
n.trees = seq(from = 100, to = 10000, by = 100)
predmat = predict(boost.boston, newdata = boston[-train,], n.trees = n.trees)
dim(predmat)

#Visualize Boosting Error Plot
boost.err = with(boston[-train,], apply( (predmat - medv)^2, 2, mean) )
plot(n.trees, boost.err, pch = 23, ylab = "Mean Squared Error", xlab = "# Trees", main = "Boosting Test Error")
abline(h = min(test.err), col = "red")

```

## Homework

1. Attempt a regression tree-based method (not covered in this tutorial) on a reasonable dataset of your choice. Explain the results. 

```{r}
library(mlbench)
data(BostonHousing)
head(BostonHousing)
BH = BostonHousing
```

CP stands for the complexity parameter and represents the benefit of adding another layer to the tree. In the table generated below, it can be observed that by the addition of the first layer increases the predictive capability of the model by 45%, and a second layer further raises its predictive probability by an additional 17%. The default cut off in R for the cp value is 0.01, and that represents the level where additional levels provide minimal benefits to improving the model.

X-error represents is another metric that can be used to measure the benefits of adding layers to the tree. X-error is the cross validation error rate, where additional layers lower the error rate. As can be observed below, additional samples decreases the error rate farily significantly until it reaches the 5th layer, where a 0.0002 decrease is observed. At that point, the rate decrease is fairly small, and additional layers will run the risk of overfitting. 

```{r}

fit <- rpart(medv~.,method="anova", data=BH)


printcp(fit) # display the results
plotcp(fit) # visualize cross-validation results



```
The below shows two plots. The first one plots the cross validation results against the actual results, whereas the second graph plots the cross validation results as a response to adding splits to the tree. 

```{r}
par(mfrow=c(1,2)) # two plots on one page
rsq.rpart(fit) # visualize cross-validation results   

```

Here I plotted the regression tree twice, once using the generic plot function, and another with the rpart plot function. The tree itself corresponds correctly with the cp and cross validation. Looking at the plotcp graph, it looks like the cut off is at around 7 splits, and that's how many splits are present in the tree below.

```{r}
plot(fit, uniform=TRUE,
   main="Regression Boston Housing Diagnosis ")
text(fit, use.n=TRUE, all=TRUE, cex=.8)

rpart.plot(fit)

```


2. Attempt both a bagging and boosting method on a reasonable dataset of your choice. Explain the results.

```{r Random Forest HW}
#set seed for reproducibility 
set.seed(30)

#split into train and test sets (300 and 206 respectively)
train = sample(1:nrow(BH), 300)

#fit training subset of data to model 
rf.bh = randomForest(medv~., data = BH, subset = train)
rf.bh

#summary of rf.boston gives information about the number of trees, the mean squared residuals (MSR), and the percentage of variance explained

#No. of variables tried at each split: 4 
#Each time the tree comes to split a node, 4 variables would be selected at random, then the split would be confined to 1 of those 4 variables.

##Lets try a range of mtry (number of variables selected at random at each split)

oob.err = double(13)
test.err = double(13)

#In a loop of mtry from 1 to 13, you first fit the randomForest to the train dataset
for(mtry in 1:13){
  fit = randomForest(medv~., data = BH, subset=train, mtry=mtry, ntree = 350)
  oob.err[mtry] = fit$mse[350] ##extract Mean-squared-error 
  pred = predict(fit, BH[-train,]) #predict on test dataset
  test.err[mtry] = with(BH[-train,], mean( (medv-pred)^2 )) #compute test error
}

#Visualize 
matplot(1:mtry, cbind(test.err, oob.err), pch = 23, col = c("red", "blue"), type = "b", ylab="Mean Squared Error")
legend("topright", legend = c("OOB", "Test"), pch = 23, col = c("red", "blue"))
```


At a glance, it looks like the model does quite well with the OOB samples where the MSE values hovering at around 5-10. The tree performs worse for the test data were the MSE values now hoever at around 15. This suggests that the model might be overfitted to the training data, where it may perform admirably on the training set, it becomes less capable of making accurate predictions on the testing data. The graph shows that the mean square error hits a low at around 4 trees, suggesting that the model performs best with that amount of trees.



```{r Boosting HW}
set.seed(93)
#Gradient Boosting Model
boost.BH = gbm(rm~., data = BH[train,], distribution = "gaussian", n.trees = 10000, shrinkage = 0.01, interaction.depth = 4)

#Variable Importance Plot
summary(boost.BH)


#Visualize important variables of interest
plot(boost.BH,i="dis")
plot(boost.BH,i="lstat")

#Predict on test set
n.trees = seq(from = 100, to = 10000, by = 100)
predmat = predict(boost.BH, newdata = BH[-train,], n.trees = n.trees)
dim(predmat)

#Visualize Boosting Error Plot
boost.err = with(BH[-train,], apply( (predmat - rm)^2, 2, mean) )
plot(n.trees, boost.err, pch = 23, ylab = "Mean Squared Error", xlab = "# Trees", main = "Boosting Test Error")
abline(h = min(test.err), col = "red")

```
The boosting summary showed that lstat and dis were the two most imformative variables in predicting the amount of rooms per dwelling. Looking online lstat stands for the lower status of the population, and medv stands for the median value of the houes. The two prediction maps also makes some intuitive sense. Houses have a higher value normally corresponds to more area - leading to more average rooms. The second prediction plot shows that as the lower status of the population increases, the number of average rooms decrease. This could be due to the fact that lower status of the population may tend to cluster around the city - an area where the average dwellings are apartments, and therefore have less rooms than the houses in the suburbs. 

A look at the boosting test error graph shows that the graph is a little scattered. Before 1000 trees the MSE dips at around 100-300 trees. There is a second visible dip in the line  at around 1600 trees. In order to optimize this model, it would be therefore recommended to run somewhere between those one of those ranges. Starting with a higher amount of trees might be a good starting place to determine just how many trees need to be run to optimize the model, as the second dip correponds to a slightly lower MSE, it may be a good starting point.
